---
title: "index"
format: html
editor: visual
---

## Setup

```{r}
library(tidyverse)
source("R/functions.R")

# Create image folder if it does not already exist
dir.create("img", showWarnings = FALSE)

wcmatches  <- read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2022/2022-11-29/wcmatches.csv')
worldcups  <- read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2022/2022-11-29/worldcups.csv')
```

## Inspect Both Tables

Each row in `wcmatches` represents one World Cup match, while each row in `worldcups` represents one World Cup tournament.

```{r}
glimpse(wcmatches)
glimpse(worldcups)

# How many tournaments? How many matches?
worldcups %>% count()
wcmatches %>% count()

# What years are covered?
worldcups %>% summarise(min(year), max(year))

# What stages exist?
wcmatches %>% count(stage)
```

## Which Countries Have Dominated?

```{r}
titles <- count_titles(worldcups)

titles
```

```{r}
titles_plot <- plot_titles(titles)

titles_plot

ggsave(
  "img/titles-by-country.png",
  plot = titles_plot,
  width = 8,
  height = 5,
  dpi = 300
)
```

Brazil has won the most World Cups with five titles, followed by Germany and Italy with four each. These three countries account for 13 of the 21 World Cups in the dataset, showing that a small number of countries have dominated the tournament historically.

## Does Hosting Help?

```{r}
wc_joined <- join_match_context(wcmatches, worldcups)

glimpse(wc_joined)
```

```{r}
host_matches <- flag_host_matches(wc_joined)

team_matches <- create_team_matches(wc_joined)

host_summary <- team_matches %>%
  group_by(is_host) %>%
  summarise(
    games = n(),
    wins = sum(won),
    win_rate = mean(won)
  )

host_summary
```

```{r}
host_win_plot <- plot_host_win_rates(host_summary)

host_win_plot
```

Host nations won 68.75% of their matches, suggesting that hosting may provide an advantage. The analyses below test whether this difference is statistically significant and whether host countries also outperform their own results from World Cups played elsewhere.

### Testing the Host Advantage

```{r}
host_test_data <- team_matches %>%
  group_by(is_host) %>%
  summarise(
    wins = sum(won),
    games = n()
  )

prop.test(
  x = host_test_data$wins,
  n = host_test_data$games
)
```

Host nations won 68.75% of their matches compared with 39.03% for non-host teams. A two-proportion test found that this difference was statistically significant (p \< 0.001), providing evidence that host nations have performed better at the World Cup.

### Accounting for Team Strength

```{r}
team_strength <- team_matches %>%
  filter(!is_host) %>%
  group_by(team) %>%
  summarise(
    historical_win_rate = mean(won),
    .groups = "drop"
  )
team_matches <- team_matches %>%
  left_join(team_strength, by = "team")

model <- glm(
  won ~ is_host + historical_win_rate,
  data = team_matches,
  family = binomial
)

summary(model)
```

Host nations won 68.75% of their matches compared with 39.03% for non-host teams. This difference was statistically significant (p \< 0.001). To determine whether this was simply because host countries tend to be stronger teams, a logistic regression was used to account for each country's historical win rate. Hosting remained statistically significant (p \< 0.001), with host teams having about 3.1 times the odds of winning compared with non-host teams with the same historical win rate. This suggests that hosting itself is associated with a meaningful advantage at the World Cup.

### Host Advantage Over Time

```{r}
host_by_year <- host_performance_by_year(team_matches)

host_by_year
```

```{r}
host_advantage_plot <- plot_host_advantage_over_time(host_by_year)

host_advantage_plot

ggsave(
  "img/host-advantage-over-time.png",
  plot = host_advantage_plot,
  width = 8,
  height = 5,
  dpi = 300
)
```

Host win rates have varied considerably across World Cups, with no clear upward or downward trend over time. The only host team with a 100% win rate in the modern era is France in 1998 which is one of the strongest teams in history, so some of the variation in host performance may be explained by the strength of the host nation rather than hosting alone.

### Do Host Nations Advance Farther?

Winning individual matches is one way to measure host advantage, but tournament progress provides another view. This analysis compares how often host and non-host teams reach the quarterfinals, semifinals, and final.

```{r}
tournament_progress <- summarize_tournament_progress(team_matches)

host_advancement <- summarize_host_advancement(tournament_progress)

host_advancement
```

```{r}
host_advancement_plot <- plot_host_advancement(host_advancement)

host_advancement_plot

ggsave(
  "img/host-tournament-advancement.png",
  plot = host_advancement_plot,
  width = 8,
  height = 5,
  dpi = 300
)
```

This comparison shows whether the host advantage extends beyond individual match wins to deeper tournament runs. If host nations reach the later rounds at higher rates, it provides additional evidence that hosting is associated with stronger overall World Cup performance.

### Do Countries Perform Better When They Host?

The earlier host-versus-non-host comparison includes many different countries, so part of the difference could come from historically strong teams being selected as hosts. A stronger comparison is to look only at countries that have hosted and compare their match win rate while hosting with their own win rate when playing in other World Cups.

```{r}
host_self_comparison <- compare_host_to_own_nonhost(team_matches)

host_self_comparison
```

```{r}
host_self_plot <- plot_host_vs_own_nonhost(host_self_comparison)

host_self_plot

ggsave(
  "img/host-vs-own-nonhost.png",
  plot = host_self_plot,
  width = 8,
  height = 6,
  dpi = 300
)
```

Host countries generally performed better when hosting the World Cup than when competing abroad. For most countries, the hosting win rate was higher than their non-hosting win rate, with particularly large increases for countries such as Uruguay, England, Sweden, and South Korea. This provides stronger evidence of a host advantage because each country is being compared with its own historical performance rather than with completely different teams. However, hosting did not improve performance for every country, suggesting that home-field advantage helps but does not guarantee success.
# Conclusion

Overall, the results suggest that hosting the World Cup is associated with a meaningful competitive advantage. Host nations won 68.75% of their matches compared with 39.03% for non-host teams, and this difference was statistically significant. Even after accounting for historical team strength, host nations still had about 3.1 times the odds of winning. However, the size of the host advantage has varied considerably between tournaments, with no clear trend over time.

These results suggest that team strength alone does not fully explain why hosts have historically performed well. Factors such as home crowd support, familiar playing conditions, reduced travel, and other advantages of playing at home could help explain the difference. However, this analysis shows an association rather than causation, and future analysis could account for factors such as opponent strength and tournament format.
